---
title: "Exploring renter population vulnerable to COVID-19"
author: "Cecile Murray"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document:
    code_folding: hide
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r setup, results='hide', message=FALSE, error=FALSE}
libs <- c("here",
          "tidyverse",
          "magrittr",
          "purrr",
          "knitr", 
          "kableExtra",
          "janitor",
          "ipumsr",
          "haven")
lapply(libs, library, character.only = TRUE)

load("../data/acs/prepared_data_2018.Rdata")
load("../data/acs/renters_2018.Rdata")

source("theme.R")
```

#### **Preliminary steps**

- Group individuals into categories by age

- Recode race/Hispanic origin variable

- Create flag for poverty, groups for % of poverty

- Group individuals into categories by household income

Technical notes:

- Renters (OWNERSHP == 2 ) excludes group quarters 


## **National numbers**

#### **Age**

- About half the people in renter households are under age 30. 

- One-quarter of people living in renter households are over 60.

```{r natl_age}
# function to compute # of people, renters, and shares of renters
get_renter_age_dist <- function(df, subdf) {
  age_table <- full_join(
    count(data, age_cat, wt = PERWT, sort = TRUE),
    count(renters, age_cat, wt = PERWT, sort = TRUE),
    by = "age_cat",
    suffix = c("_all", "_renters")
  ) %>% 
    mutate(share_renting = n_renters / n_all,
           share_of_renters = n_renters / sum(n_renters),
           cumulative_share_of_renters = cumsum(share_of_renters))
  return(age_table)
}

natl_age_table <- get_renter_age_dist(data, renters)

kable(natl_age_table,
      caption = "Age distribution of renters, 2018",
      col.names = c("Age", "Total #", "# in renter households", "Share renting",
                    "Share of renters", "Cumulative share of renters"),
      digits = 2) %>% 
  kable_styling(latex_options = "HOLD_position")
```

#### **Poverty status**

```{r renter_poverty}
# summarize poverty for people in renter households
get_pov_by_age_dist <- function(df) {
  rv <- df %>% 
    group_by(poor) %>% 
    count(age_cat, wt = PERWT, sort = TRUE) %>% 
    spread("poor", "n") %>% 
    clean_names() %>% 
    dplyr::rename("not_poor" = "false",
                  "poor" = "true") %>% 
    mutate(total = poor + not_poor,
           povrate = poor / total)
  return(rv)
}

natl_renter_poverty <- get_pov_by_age_dist(renters)

# number of poor people in renter households by age
natl_renter_poverty %>% 
  ggplot(aes(x = reorder(age_cat, -total),
             y = poor)) +
  geom_col(fill = terner_blue) +
  geom_text(aes(label = scales::comma(poor)),
            # family = "Minion Pro",
            nudge_y = 300000) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "More poor renters are young",
       subtitle = "Poor people living in renter households by age category, 2018",
       x = "Age Category",
       y = "Number of poor people",
       caption = "Source: ACS PUMS via IPUMS") +
  terner_theme()


natl_renter_poverty %>% 
  ggplot(aes(x = reorder(age_cat, -total),
             y = povrate)) +
  geom_col(fill = terner_blue) +
  geom_text(aes(label = round(povrate * 100, 2)),
            # family = "Minion Pro",
            nudge_y = .01) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "About a fifth of older renters are poor",
       subtitle = "Share of people living in renter households by age category, 2018",
       x = "Age Category",
       y = "Share in poverty",
       caption = "Source: ACS PUMS via IPUMS") +
  terner_theme()

```

```{r renter_poverty_level}
# summarize poverty for people in renter households
natl_renter_pov_by_lvl <- renters %>%
  group_by(pov_level) %>% 
  count(age_cat, wt = PERWT, sort = TRUE) %>% 
  ungroup() %>% 
  group_by(pov_level) %>% 
  mutate(n_pov = sum(n),
         age_share = n / n_pov) 

                          
# number of poor people in renter households by age
natl_renter_pov_by_lvl %>% 
  ggplot(aes(x = age_cat,
             y = age_share)) +
  geom_col(position = "dodge",
           fill = terner_blue) +
  facet_wrap(~pov_level) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Poorer renters skew younger",
       subtitle = "Share of people of each age living in renter households by % of poverty, 2018",
       x = "Age Category",
       y = "Percent of people in renter households",
       caption = "Source: ACS PUMS via IPUMS") +
  terner_theme(axis.text.x = element_text(angle = 45))
```

#### **Income**

```{r natl_income}
# summarize renters by income category
natl_renter_income <- renters %>% 
  count(income_level, wt = PERWT) %>% 
  mutate(income_share = n / sum(n))

# plot the numbers
natl_renter_income %>% 
  ggplot(aes(x = income_level,
             y = n)) +
  geom_col(fill = terner_blue) +
  geom_text(aes(label = scales::comma(n)),
            # family = "Minion Pro",
            nudge_y = 1000000) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Most renters have low or modest incomes",
       subtitle = "Number of people in renter households by household income, 2018",
       x = "Household income bracket",
       y = "",
       caption = "ACS 2018 via IPUMS") +
  terner_theme()

# plot the percentages
natl_renter_income %>% 
  ggplot(aes(x = income_level,
             y = income_share)) +
  geom_col(fill = terner_blue) +
  geom_text(aes(label = scales::percent(income_share)),
            # family = "Minion Pro",
            nudge_y = 0.01) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Most renters have low or modest incomes",
       subtitle = "Share of people in renter households by household income, 2018",
       x = "Household income bracket",
       y = "",
       caption = "ACS 2018 via IPUMS") +
  terner_theme()

```

#### **Cost burden**

```{r natl_cost_burden}
natl_cost_by_age <-  renters %>% 
  mutate(cost_burden = RENT / HHINCOME,
         burden_status = case_when(
           HHINCOME == 0 ~ "Zero income",
           cost_burden < 0.3 ~ "Not burdened",
           between(cost_burden, 0.3, 0.5) ~ "30-50% cost burdened",
           between(cost_burden, 0.500001, 1) ~ "50-100% cost burdened",
           cost_burden > 1 ~ "> 100% cost burdened"
         ),
         burden_status = factor(burden_status,
         levels = c("Zero income", "> 100% cost burdened",
                    "50-100% cost burdened",
                    "30-50% cost burdened",
                    "Not burdened"))) %>% 
  group_by(age_cat) %>% 
  count(burden_status, wt = PERWT) %>% 
  group_by(burden_status) %>% 
  mutate(n_age = sum(n),
            age_share = n / n_age)
  
natl_cost_by_age %>% 
  ggplot(aes(x = burden_status,
             fill = age_cat)) +
  geom_col(aes(y = age_share)) +
  scale_y_continuous(labels = scales::percent) +
  coord_flip() +
  labs(title = "Cost-burdened renters skew younger",
       subtitle = "Share of individuals of each age by gross rent as a share of household income, 2018",
       x = "",
       y = "% of people in renter households",
       fill = "Age group") +
  terner_theme()


```


### **State-by-state numbers**

```{r state_old_renters}
state_old_renters <- 
  renters %>% 
  group_by(STATEFIP) %>% 
  count(age_cat, wt = PERWT) %>% 
  mutate(n_state = sum(n),
         age_share = n / n_state) %>% 
  ungroup() %>% 
  mutate(STATEFIP = str_pad(STATEFIP, 2, side = "left", pad = "0")) %>% 
  left_join(
    distinct(select(tidycensus::fips_codes, state_code, state)),
    by = c("STATEFIP" = "state_code"))

OLDER <- c("50-59", "60-69", "70-79", "80-89", "90-99", "100+")
natl_older_share <- sum(natl_age_table$share_of_renters[natl_age_table$age_cat %in% OLDER])

state_old_renters %>% 
  filter(age_cat %in% OLDER) %>% 
  group_by(state) %>% 
  summarize(age_share = sum(age_share)) %>% 
  ggplot(aes(x = reorder(state, age_share),
             y = age_share)) +
  geom_col(fill = terner_blue) +
  geom_hline(yintercept = natl_older_share) +
  scale_y_continuous(label = scales::percent) +
  labs(title = "A larger share of renters in the Northeast are older",
    subtitle = "Share of individals aged 50+ living in renter households by state, 2018",
       x = "", 
       y = "Share") +
  annotate(geom = "text",
           x = 7, y = natl_older_share + .01,
           label = "National share: 22%") +
  terner_theme(axis.text.x = element_text(angle = 45, size = 9))
```


#### **Occupations**

- need to sort based on ability to telework

#### Additional ideas

- type of health insurance coverage: employer vs state sponsored

- look for multigenerational households, perhaps by # of rooms/bedrooms


### **Metro area numbers**

- TO DO: check PUMA-MSA crosswalk