---
title: "Exploring renter population vulnerable to COVID-19"
author: "Cecile Murray"
date: "3/23/2020"
output: 
  html_document:
    code_folding: hide

---

### **Setup**

```{r setup, results='hide', message=FALSE, error=FALSE}
libs <- c("here",
          "tidyverse",
          "magrittr",
          "purrr",
          "knitr", 
          "kableExtra",
          "janitor",
          "ipumsr",
          "haven")
lapply(libs, library, character.only = TRUE)

load("../data/acs/acs_2018.Rdata")
```

#### **Recoding**

- Group individuals into categories by age

- Recode race/Hispanic origin variable

- Create flag for poverty

```{r recode, warning=FALSE, message=FALSE}
# generate age categories, race/ethnicity recode
data %<>% 
  mutate(
    age_cat = case_when(
      AGE < 20 ~ "Under20",
      between(AGE, 20, 29) ~ "20-29",
      between(AGE, 30, 39) ~ "30-39",
      between(AGE, 40, 49) ~ "40-49",
      between(AGE, 50, 59) ~ "50-59",
      between(AGE, 60, 69) ~ "60-69",
      between(AGE, 70, 79) ~ "70-79",
      between(AGE, 80, 89) ~ "80-89",
      between(AGE, 90, 99) ~ "90-99",
      AGE >= 100 ~ "100+"),
    raceth = case_when(
      RACE == 1 & HISPAN %in% c(0, 9) ~ "white",
      RACE == 2 & HISPAN %in% c(0, 9) ~ "black",
      RACE == 3 & HISPAN %in% c(0, 9) ~ "ai_an",
      between(RACE, 4, 6) & HISPAN %in% c(0, 9) ~ "east_asian",
      RACE > 6 & HISPAN %in% c(0, 9) ~ "other_mult",
      between(HISPAN, 1, 4) ~ "hispanic",
      TRUE ~ "not determined"
    ),
    poor = if_else(POVERTY < 100, TRUE, FALSE)) 

# filter to only renters
renters <- data %>% filter(OWNERSHP == 2)
```

### **National numbers**

#### Age

About half the people in renter households are under age 30. Only one-quarter of these people are over 60.

```{r natl_age}
# function to compute # of people, renters, and shares of renters
get_renter_age_dist <- function(df, subdf) {
  age_table <- full_join(
    count(data, age_cat, wt = PERWT, sort = TRUE),
    count(renters, age_cat, wt = PERWT, sort = TRUE),
    by = "age_cat",
    suffix = c("_all", "_renters")
  ) %>% 
    mutate(share_renting = n_renters / n_all,
           share_of_renters = n_renters / sum(n_renters),
           cumulative_share_of_renters = cumsum(share_of_renters))
  return(age_table)
}

natl_age_table <- get_renter_age_dist(data, renters)

kable(natl_age_table,
      caption = "Age distribution of renters, 2018")
```

**Poverty status**

```{r renter_poverty}
# summarize poverty for people in renter households
get_pov_by_age_dist <- function(df) {
  rv <- df %>% 
    group_by(poor) %>% 
    count(age_cat, wt = PERWT, sort = TRUE) %>% 
    spread("poor", "n") %>% 
    clean_names() %>% 
    dplyr::rename("not_poor" = "false",
                  "poor" = "true") %>% 
    mutate(total = poor + not_poor,
           povrate = poor / total)
  return(rv)
}

natl_renter_poverty <- get_pov_by_age_dist(renters)

# number of poor people in renter households by age
natl_renter_poverty %>% 
  ggplot(aes(x = reorder(age_cat, -total),
             y = poor)) +
  geom_col() +
  geom_text(aes(label = poor), nudge_y = 300000) +
  scale_y_continuous(labels = scales::number) +
  labs(title = "More poor renters are young",
       subtitle = "Poor people living in renter households by age category, 2018",
       x = "Age Category",
       y = "Number of poor people",
       caption = "Source: ACS PUMS via IPUMS") +
  theme_minimal()


natl_renter_poverty %>% 
  ggplot(aes(x = reorder(age_cat, -total),
             y = povrate)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "About a fifth of older renters are poor",
       subtitle = "Share of people living in renter households by age category, 2018",
       x = "Age Category",
       y = "Share in poverty",
       caption = "Source: ACS PUMS via IPUMS") +
  theme_minimal()

```

```{r renter_poverty_nokids}
# summarize poverty for people in renter households, no kids/teens
natl_renter_poverty_nokids <- renters %>% 
  filter(AGE > 19) %>% 
  get_pov_by_age_dist()

# number of poor people in renter households by age
natl_renter_poverty_nokids %>% 
  ggplot(aes(x = reorder(age_cat, -total),
             y = poor)) +
  geom_col() +
  geom_text(aes(label = poor), nudge_y = 200000) +
  scale_y_continuous(labels = scales::number) +
  labs(title = "More poor renters are young",
       subtitle = "Poor people living in renter households by age category, 2018",
       x = "Age Category",
       y = "Number of poor people",
       caption = "Source: ACS PUMS via IPUMS") +
  theme_minimal()


natl_renter_poverty_nokids %>% 
  ggplot(aes(x = reorder(age_cat, -total),
             y = povrate)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "About a fifth of older renters are poor",
       subtitle = "Share of people living in renter households by age category, 2018",
       x = "Age Category",
       y = "Share in poverty",
       caption = "Source: ACS PUMS via IPUMS") +
  theme_minimal()
```

#### **Income**

- occupations: need to sort based on ability to telework

- health insurance coverage?

- look for multigenerational households, perhaps by # of rooms/bedrooms


Technical notes:

- this definition of renters excludes group quarters (OWNERSHP == 2 )

### **State-by-state numbers**


### **Metro area numbers**

- TO DO: build PUMA-MSA crosswalk